The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2016-03-07 oil_prices_obes.tex
minor improvements)

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
Words: 7000
\end{abstract}

@@

A large body of economic litterature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry. This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.

@@

By looking only at the effect of price on fields that currently or previously have produced oil, I am limiting the scope of this article.  The effect of oil prices on total production over an extended period of time is due not just to reactions in production in existing fields but also increased searching for new fields.  In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as production from previously un-economic fields.

@@

\section{Econometric Litterature on The Effect of Oil Price on Production}

@@

An extensive litterature exists that attempts to estimate the effect of oil prices on regional or global oil supply. Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level. \citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply. \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an emperical model of oil production in the US. The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production. The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.

@@

\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s. Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of litterature.

The economics litterature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted. \citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time. Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown.

@@

On the surface, this article fits well into the previous econometrics litterature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces. However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production. This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields. While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.

@@

Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.  Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.  The central government also receives revenues through its majority ownership stake in Statoil. In addition, a wholly state-owned company, Petero, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).

@@

\section{A Generalized Additive Model of Oil Field Production}
Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis. However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.

@@

For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}. In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2st to 4th lags, with a peak at the 2nd lag. A negative coefficient for the eighth lag is estimated in the 2-d model, however this is not significant in the model with random effects. Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model. On the other hand, the block of terms for the 1st through 4th lags do significantly add to the fit of the model.

@@

To attempt to capture this heterogeneity, I introduce an indicator variable that represents the build-out phase defined as the period between the start of production up to peak production. I add interaction effects between this indicator variable and the block of significant lagged price effects $price\_l1$ through $price\_l4$.

@@

The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}. The interaction effects, labeled $price\_lt:build$, are not estimated to be significantly different from zero. Given the small sample used to estimate these terms, this is not surprising. On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now subsantially larger in magnitude. In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase. This provides some evidence that the effect of price on production is indeed strongest in this latter stage of field development.

@@

Policy implications arise for both energy industry regulators and financial policy makers. Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas. This is particularily important on the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices. \footnote{I thank an anonymous reviewer for pointing out this implication.}

>>>>

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production. I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
Words: 7100
\end{abstract}

@@

A large body of economic literature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry. This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.

@@

By looking only at the effect of price on production in fields that currently or previously have produced oil, I am limiting the scope of this article.  The effect of oil prices on total production over the long term is due to both the reaction of production in existing fields as well as searching for new fields. This article is limited to the decisions oil companies can make to increase production in existing fields like making use of spare capacity or increasing production drilling. In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as development of previously un-economic fields.

@@

\section{Econometric Literature on The Effect of Oil Price on Production}

@@

An extensive literature exists that attempts to estimate the effect of oil prices on regional or global oil supply. Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level. \citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply. \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an empirical model of oil production in the US. The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production. The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.

@@

\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s. Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of literature.

The economics literature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted. \citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time. Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown.

@@

On the surface, this article fits well into the previous econometrics literature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces. However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production. This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields. While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.

@@

Norwegian waters in the Barents Sea have also been open to exploration since the 1980's, however up until recently only a few, small finds were made and none came into commercial production.  Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.  The central government also receives revenues through its majority ownership stake in Statoil. In addition, a wholly state-owned company, Petoro, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).

@@

\section{A Generalized Additive Model of Oil Field Production}
Parametric linear models have the sizable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis. However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.

@@

For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}. In general the results from the two models agree with each other, indicating little to no effect on the concurrent price term, but with positive coefficients estimated on the 2nd to 4th lags, with a peak at the 2nd lag. A negative coefficient for the eighth lag is estimated in the 2-d model, however this is not significant in the model with random effects. Furthermore, F-tests with comparisons to restricted models without the concurrent and 5-8th lags indicate that these terms do not as a group significantly improve the fit of the model. On the other hand, the block of terms for the 1st through 4th lags do significantly add to the fit of the model.

@@

To attempt to capture this heterogeneity, I introduce an indicator variable that represents the build-out phase defined as the period between the start of production up to peak production. I add interaction effects between this indicator variable and the block of lagged price effects $price\_l1$ through $price\_l4$.

@@

The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}. The interaction effects, labeled $price\_lt:build$, are not estimated to be significantly different from zero. Given the small sample used to estimate these terms, this is not surprising. On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now substantially larger in magnitude. In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase. This provides some evidence that the effect of price on production is indeed strongest in this latter stage of field development.

@@

Policy implications arise for both energy industry regulators and financial policy makers. Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas. This is particularly important on the Norwegian Continental Shelf, as several new fields have recently come on-line following a period of high oil prices. \footnote{I thank an anonymous reviewer for pointing out this implication.}
